# Practice Problems: Introductory Example {.unnumbered}

{{< include _r_setup.qmd >}}

{{< include _python_setup.qmd >}}





Suppose we want to estimate the proportion of Cal Poly students that are left handed, denoted $\theta$.
Suppose first that only the values 0.05, 0.10, 0.15, 0.20, 0.25 are plausible.

1. Create a table and plot of your prior plausibility assessment, like we did in Example 2.1.
Assume that each of the five values has some plausibility, no matter how small; don’t assign 0 plausibility to any of the values.
There is no one right answer; just explain your reasoning.

1. Suppose we'll collect data on a sample of 30 Cal Poly students; let $y$ be the number of students in the sample who are left handed.
Explain in words how you could use physical objects (dice, spinners, etc) to simulate a $(\theta, y)$ pair, in a way that reflects your plausibility assessment from the previous part.

1. Now suppose you collect data on a random sample of 30 Cal Poly students and 4 are left-handed.
Explain in words how you could use simulation to assess your plausibility of the five possible values of $\theta$ after observing sample data.
In particular, describe in words how you would use the simulation to approximate what percent of your total plausibility the value 0.10 accounts for after observing the sample data.
(To be clear: you don't have to actually carry out the simulation or write code. You just need to describe the process in words.)

1. Which of the values 0.05, 0.10, 0.15, 0.20, 0.25 do you think are more plausible after observing the sample data (4/30) than before?
Less plausible after than before?
Explain intuitively, without actually running the simulation or computing the posterior plausibility.

## Solution

(@) A quick online search reveals that somewhere around 10% of people in general are left-handed.
There is no reason to assume Cal Poly would be different, so we'll assume 0.10 is most plausible and 0.25 is least plausible, say:

- 0.25 is the least plausible value
- 0.20 is 5 times more plausible than 0.25
- 0.15 is 2.4 times more plausible than 0.20
- 0.05 is as plausible as 0.15
- 0.10 is 5/3 times more plausible than 0.05.

```{r}
theta = seq(0.05, 0.25, by = 0.05)

prior_units = c(2.4 * 5, 5/3 * 2.4 * 5, 2.4 * 5, 5, 1)

prior = prior_units / sum(prior_units)

data.frame(theta, prior_units, prior) |>
  adorn_totals("row") |>
  kbl() |> kable_styling()
```

(@) Stage 1: Create a spinner corresponding to your prior plausibility: it lands on 0.10 with probability 0.40, ..., 0.25 with probability 0.02.
Spin the spinner to simulate $\theta$.

    Stage 2: Given $\theta$, create a spinner that lands on "success" with probability $\theta$, then spin it 30 times and count the number of spins that land on "success" to simulate $y$.
For example, if $\theta = 0.05$, create a spinner than lands on "success" with probability 0.05 then spin it 30 times and count the number of spins that land on "success".

(@) First, repeat the two-stage simulation in the previous part to simulate many $(\theta, y)$ pairs.
Then discard $(\theta, y)$ pairs for which $y\neq 4$.
Among the remaining $(\theta, y)$ pairs (all with $y=4$), summarize the relative frequencies of the $\theta$ values.
These conditional relative frequencies approximate the posterior plausibility.
For example, to approximate the posterior plausibility that $\theta$ is 0.10, among the non-discarded repetitions count the number of repetitions with $\theta = 0.1$ and divide by the total number of non-discarded repetitions.

(@) The observed sample proportion is $4/30=0.133$.
So plausibility probably increases for 0.10 and 0.15 and decreases for 0.25.
The values 0.05 and 0.15 started off as equally plausible, but after observing sample data 0.15 will be more plausible than 0.05.



```{r}
n_rep = 100000

theta_sim = sample(theta, n_rep, replace = TRUE, prob = prior)

n = 30

y_sim = rbinom(n_rep, n, theta_sim)

sim = data.frame(theta_sim, y_sim)

y_obs = 4

sim_conditional = sim |>
  filter(y_sim == y_obs)

sim_conditional |>
  count(theta_sim, name = "freq") |>
  mutate(rel_freq = freq / sum(freq)) |>
  adorn_totals("row") |>
  kbl(digits = 4) |> kable_styling()

```

