# Interpretations of Probability and Statistics {#interpretations}

You have some familiarity with "probability" or "chance" or "odds".  But what do we really mean when talk about "probability"?  It turns out there are two basic interpretations: relative frequency and "subjective" probability.  These two interpretations provide the philosophical foundation for two schools of statistics: frequentist (hypothesis tests and confidence intervals that you've seen before) and Bayesian (which this course is about).  This chapter introduces the two interpretations.





## Instances of randomness {#randomness}

Probability comes up in a wide variety of situations. Consider just a few examples.

1. The probability that a single flip of a fair coin lands on heads.
1. The probability you win the next [Powerball lottery](https://www.powerball.com/) if you purchase a single ticket, 6-7-16-23-26, plus the Powerball number, 4.
1. The probability that a “randomly selected” Cal Poly student is a California resident.
1. The probability that it rains tomorrow in San Luis Obispo.
1. The probability that there are more Atlantic hurricanes in 2021 than in 2020.
1. The probability that the Green Bay Packers win the next Superbowl.
1. The probability that Democrats win both Senate seats in the Georgia runoff election.
1. The probability that extraterrestrial life currently exists somewhere in the universe.
1. The probability that Alexander Hamilton actually wrote [51](https://www.youtube.com/watch?v=DPgE7PNzXag) of the [Federalist Papers](https://en.wikipedia.org/wiki/The_Federalist_Papers). (The papers were published under a common pseudonym and authorship of some of the papers is disputed.)
1. The probability that you ate an apple on April 17, 2009.


```{example randomness}

How are the situations above similar, and how are they different?  What is one feature that all of the situations have in common? Is the interpretation of "probability" the same in all situations?  Take some time to consider these questions before looking at the solution. The goal here is to just think about these questions, and not to compute any probabilities (or to even think about how you would).

```

```{solution randomness-sol}
to Example \@ref(exm:randomness)
```

```{asis, fold.chunk = TRUE}

This exercise is intended to motivate discussion, so you might have thought of some other ideas we don't address here.  That's good! But here are a few thoughts we specifically want to mention now.

The one feature that all of the situations have in common is *uncertainty*.  Sometimes the uncertainty arises from a repeatedable physical phenomenon that can result in multiple potential outcomes, like flipping a coin or drawing the winning Powerball number.  In other cases, there is uncertainty because the probability concerns the future, like tomorrow's high temperature or the result of the next Superbowl.  But there can also be uncertainty about the past: there are some Federalist papers for which the author is unknown, and you probably don't know for sure whether or not you ate an apple on April 17, 2009.

Whenever there is uncertainty, it is reasonable to consider relative likelihoods of potential outcomes.  For example, even though you don't know for certain whether you ate an apple on April 17, 2009, if you're usually an apple-a-day person you might think the probability is high.  We don't know for sure what team will win the next Superbowl, but we might think that the Packers are more likely than the Las Vegas (?!?!) Raiders to be the winner.

While all of the situations involve uncertainty, it seems that there are different "types" of uncertainty.  Even though we don't know if a coin flip will land on heads or tails, the notion of "fairness" implies that these two outcomes are "equally likely".   Likewise, there are some rules to how the Powerball drawing works, and it seems like these rules should determine the probability of drawing that particular winning number.

However, there aren't any specific "rules of uncertainty" that govern whether or not you ate an apple on April 17, 2009.  You either did or you didn't, but that doesn't mean the two outcomes are necessarily equally likely.  Regarding the Superbowl, of course there are rules that govern the NFL season and playoffs, but there are no "rules of uncertainty" that tell us with certainty which teams will win individual games. 


It also seems that there are different interpretations of probability. Given that a coin is fair, we might all agree that the probability that it lands on heads is 1/2.  Similarly, given the rules of the Powerball lottery, we might all agree on the probability that a drawing results in a particular winning number.  However, there isn't necessarily consensus about the probability that it rains tomorrow in San Luis Obispo.  Different weather forecasters, news stations, or websites might provide different values for this probability.  There seems to be some subjectivity to probability in situations like tomorrow's weather or the next Superbowl. 

Finally, some of these phenomenon are repeatedable. We could (in principle) flip a coin many times and how often the flips land on heads, or repeat the Powerball drawing over and over to see how the winning numbers behave. However, many of these situations involve something that only happens once, like tomorrow or April, 17, 2009 or the next Superbowl. Even when the phenomenon happens only once it reality, we can still develop models of what might happen if we were to hypothetically repeat the phenomenon many times.   For example, meteorologists use historical data and meteorological models to forecast [potential paths of a hurricane](https://www.nhc.noaa.gov/cone_usage.php).



```
	




The subject of probability concerns *random* phenomena. A phenomenon is **random** if there are multiple potential outcomes, and there is **uncertainty** about which outcome will occur. Uncertainty is understood in broad terms, and in particular does not only concern future occurrences.

Some phenomena involve physical randomness^[We will refer to as "random" any scenario that involves a reasonable degree of uncertainty.  We're avoiding philosophical questions about what is "true" randomness, like the following.  Is a coin flip really random? If all factors that affect the trajectory of the coin were known precisely, then wouldn't the outcome be determined?  Does true randomness only exist in quantum mechanics?], like flipping a coin or drawing powerballs at random from a bin.  In many other situations randomness just vaguely reflects uncertainty.

Contrary to colloquial uses of the word, random does *not* mean haphazard. In a random phenomenon, while
individual outcomes are uncertain, there is a *regular distribution of
outcomes over a large number of (hypothetical) repetitions*.

- In two flips of a fair coin we wouldn't necessarily see one head and one tail. But in 10000 flips of a fair coin, we might expect to see close to 5000 heads and 5000 tails. 
- We don't know who will win the next Superbowl, but we can and should consider some teams as more likely to win than others.  We could imagine a large number of hypothetical 2020-2021 seasons; how often would we expect the Eagles to win? The Raiders? (Hopefully a lot for the Eagles; probably not much for the Raiders).


Random also does *not* necessarily mean equally likely. In a random
phenomenon, certain outcomes or events might be more or less likely than
others.  

- It's much more likely than not that a randomly selected Cal Poly student is a California resident.
- Not all NFL teams are equally likely to win the next Superbowl.

Finally, randomness is also not necessarily undesirable.  In particular, many statistical applications often employ the planned use of randomness with the goal of collecting "good" data.

- *Random selection* involves selecting a sample of individuals "at random" from a population (e.g., via random digit dialing), with the goal of selecting a representative sample.  
- *Random assignment* involves assigning individuals at random to groups (e.g., in a randomized experiment), with the goal of constructing groups that are similar in all aspects.

The **probability** of an event associated with a random phenomenon is a number in the interval $[0, 1]$ measuring the event's likelihood or degree of uncertainty. A probability can take any values in the continuous scale from 0% to 100%^[Probabilities are usually defined as decimals, but are often colloquially referred to as percentages.  We're not sticklers; we'll refer to probabilities as decimals and as percentages.]. In particular, a probability requires much more interpretation than "is the probability greater than, less than, or equal to 50%?" As Example \@ref(exm:randomness) suggests, there can be different interpretations of "probability", which we'll start to explore in the next section.



## Interpretations of probability {#interpretations-of-probability}


In the previous section we encountered a variety of scenarios which involved uncertainty, a.k.a. randomness.  Just as there are a few "types" of randomness, there are a few ways of interpreting probability, namely, *long run relative frequency* and *subjective probability*.








### Long run relative frequency {#rel-freq}

We can all agree that the probability that a single flip of a fair coin lands on heads is 1/2, a.k.a., 0.5, a.k.a, 50%.  After all, the notion of "fairness" implies that the two outcomes, heads and tails, should be equally likely, so we have a "50/50 chance" of heads.  But how else can we interpret this 50%?  One interpretation involves considering *what would happen if we flipped the coin main times*.  Now, if we would flipped the coin twice, we wouldn't expect to necessarily see one head and one tail.  But in many flips, we might expect to see heads on something close to 50\% of flips.

```{r, echo = FALSE}

n = 1000
i = 1:n
x = sample(c("H", "T"), size = n, replace = TRUE)
phat_n = cumsum(x == "H") / i

```


Let's try this out. Table \@ref(tab:coin-flips) displays the results of 10 flips of a fair coin. The first column is the flip number and the second column is the result of the flip.  The third column displays the *running proportion of flips that result in H*. For example, the first flip results in `r x[1]` so the running proportion of H after 1 flip is `r sum(x[1] == "H")`/1; the first two flips result in (`r x[1]`, `r x[2]`) so the running proportion of H after 2 flips is `r sum(x[1:2] == "H")`/2; and so on. Figure \@ref(fig:coin-flips-plot) plots the running proportion of H by the number of flips.  We see that with just a small number of flips, the proportion of H fluctuates considerably and is not guaranteed to be close to 0.5.  Of course, the results depend on the particular sequence of coin flips.  We encourage you to flip a coin 10 times and compare your results.


(ref:cap-coin-flips) Results and running proportion of H for 10 flips of a fair coin.

```{r, coin-flips, echo = FALSE}

n1 = 10

knitr::kable(
  data.frame(i[1:n1], x[1:n1], phat_n[1:n1]),
  align = "r",
  col.names = c("Flip", "Result", "Running proportion of H"),
  booktabs = TRUE,
  caption = "(ref:cap-coin-flips)",
  digits = 3
)

```  



(ref:cap-coin-flips-plot) Running proportion of H versus number of flips for the 10 coin flips in Table \@ref(tab:coin-flips).


```{r coin-flips-plot, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="(ref:cap-coin-flips-plot)"}

plot(i[1:n1], phat_n[1:n1], type = "o",
     xlab = "Flip number",
     ylab = "Running proportion of H",
     ylim = c(0, 1),
     col = "blue",
     xaxt = "n")
axis(1, i[1:n1])
abline(h = 0.5, lty = 2)

```


Now we'll flip the coin 90 more times for a total of 100 flips.  The plot on the left in Figure \@ref(fig:coin-flips-plot2) summarizes the results, while the plot on the right also displays the results for 3 additional sets of 100 flips. The running proportion fluctuates considerably in the early stages, but settles down and tends to get closer to 0.5 as the number of flips increases.  However, even after 100 flips the proportion of flips that result in H isn't guaranteed to be very close to 0.5.


(ref:cap-coin-flips-plot2) Running proportion of H versus number of flips for four sets of 100 coin flips.


```{r coin-flips-plot2, echo=FALSE, warning=FALSE, message=FALSE, fig.show="hold", out.width="50%", fig.cap="(ref:cap-coin-flips-plot2)"}

n2 = 100

plot(i[1:n2], phat_n[1:n2], type = "o",
     xlab = "Flip number",
     ylab = "Running proportion of H",
     ylim = c(0, 1),
     col = "blue",
     xaxt = "n")
axis(1, seq(0, n2, n2 / 10))
abline(h = 0.5, lty = 2)

x2 = cbind(as.numeric(x == "H"),
           matrix(rbinom(3 * n, 1, 0.5), ncol = 3))
phat_n2 = sweep(apply(x2, 2, cumsum), 1, i, FUN = '/')

matplot(cbind(i, i, i, i)[1:n2, ], phat_n2[1:n2, ], type = "l",
     xlab = "Flip number",
     ylab = "Running proportion of H",
     ylim = c(0, 1),
     col = c("blue", "orange", "seagreen", "purple"),
     xaxt = "n")
axis(1, seq(0, n2, n2 / 10))
abline(h = 0.5, lty = 2)



```


Now for each set of 100 flips, we'll flip the coin 900 more times for a total of 1000 flips in each of the four sets.  The plot on the left in Figure \@ref(fig:coin-flips-plot3) summarizes the results for our original set, while the plot on the right also displays the results for the three additional sets from Figure \@ref(fig:coin-flips-plot3). Again, the running proportion fluctuates considerably in the early stages, but settles down and tends to get closer to 0.5 as the number of flips increases.  There is less variability in the proportion of H after 1000 flips than after 100.  Now, even after 1000 flips the proportion of flips that result in H isn't guaranteed to be exactly 0.5, but we see a tendency for the proportion to get closer to 0.5 as the number of flips increases.


(ref:cap-coin-flips-plot3) Running proportion of H versus number of flips for four sets of 1000 coin flips.


```{r coin-flips-plot3, echo=FALSE, warning=FALSE, message=FALSE, fig.show="hold", out.width="50%", fig.cap="(ref:cap-coin-flips-plot3)"}

n2 = 1000

plot(i[1:n2], phat_n[1:n2], type = "l",
     xlab = "Flip number",
     ylab = "Running proportion of H",
     ylim = c(0, 1),
     col = "blue",
     xaxt = "n")
axis(1, seq(0, n2, n2 / 10))
abline(h = 0.5, lty = 2)

x2 = cbind(as.numeric(x == "H"),
           matrix(rbinom(3 * n, 1, 0.5), ncol = 3))
phat_n2 = sweep(apply(x2, 2, cumsum), 1, i, FUN = '/')

matplot(cbind(i, i, i, i)[1:n2, ], phat_n2[1:n2, ], type = "l",
     xlab = "Flip number",
     ylab = "Running proportion of H",
     ylim = c(0, 1),
     col = c("blue", "orange", "seagreen", "purple"),
     xaxt = "n")
axis(1, seq(0, n2, n2 / 10))
abline(h = 0.5, lty = 2)



```

In summary, in a large number of flips of a fair coin we expect about 50% of flips to result in H.  That is, the probability that a flip of a fair coin results in H can be interpreted as the *long run proportion of flips that result in H*, or in other words, the *long run relative frequency of H*.

In general, the probability of an event associated with a random phenomenon can be interpreted as a **long run proportion** or **long run relative frequency**: the probability of the event is the proportion of times that the event would occur in a very large number of hypothetical repetitions of the random phenomenon.
<!-- A natural question is: "how many repetitions are required to represent the long run?"  We'll return to this question and the idea of long run relative frequency in later chapters. -->


The long run relative frequency interpretation of probability can be applied when a situation can be repeated numerous times, at least conceptually, and the outcome can be observed each time. One benefit of the relative frequency interpretation is that the probability of an event can be *approximated* by simulating the random phenomenon a large number of times and determining the proportion of simulated repetitions on which the event occurred out of the total number of repetitions in the simulation. A **simulation** involves an artificial recreation of the random phenomenon, usually using a computer.  After many repetitions the relative frequency of the event will settle down to a single constant value, and that value is the approximately the probability of that event.

Of course, the accuracy of simulation-based approximations of probabilities depends on how well the simulation represents the actual random phenomenon. Conducting a simulation can involve many assumptions which influence the results. Simulating many flips of a fair coin is one thing; simulating an entire NFL season and the winner of the Superbowl is an entirely different story.


```{example interpret-rel-freq}

In each of the following, write a clearly worded sentence interpreting the numerical value of the probability as a long run relative frequency in context.  (Just take the numerical values--- 0.1, 0.25, and 0.73 --- as given. We'll see how to compute probabilities later.)

1. The probability that a roll of a fair ten-sided die lands on 1 is 0.1.
1. The probability that two flips of a fair coin both land on H is 0.25.
1. The probability that in 100 flips of a fair coin the proportion of flips that land on H is between 0.45 and 0.55 is 0.73.

```


```{solution interpret-rel-freq-sol}
to Example \@ref(exm:interpret-rel-freq)
```

```{asis, fold.chunk = TRUE}

1.  About 10% of rolls of a fair ten-sided result in a roll of 1. The phenomenon is a roll of a far ten-sided die and the event of interest is whether or not the die lans on 1. 
1. In about 25% of sets of two fair coin flips, both flips in the set land on H.   The phenomenon involves *two* flips of a coin, so we consider what would happen over many *sets* of two flips each.
1. In about 73% of sets of 100 fair coin flips, the proportion of H for the set is between 0.45 and 0.55.  The phenomenon involves 100 coin flips, so we consider many sets of 100 coin flips each, each set resulting in a proportion of H that is either between 0.45 and 0.55 or not.  Imagine adding many more paths to the plot on the right in Figure \@ref(fig:coin-flips-plot2), each corresponding to a set of 100 flips, and seeing how many of the paths result in a value between 0.45 and 0.55 at flip 100.

```


### Subjective probability {#subjective-probability}

The long run relative frequency interpretation is natural in  repeatable situations like flipping a coin, drawing Powerballs from a bin, or selecting a Cal Poly student at random.

On the other hand, it is difficult to conceptualize some scenarios in the long run.  Superbowl 2021 will only be played once, the 2021 Georgia run off election will only be conducted once, and there was only one April 17, 2009 on which you either did or did not eat an apple.  But while these situations are not naturally repeatable they still involve randomness (uncertainty) and it is still reasonable to assign probabilities.  At this point in time, the Kansas City Chiefs are more likely than the [Jacksonville Jaguars](https://tenor.com/0JBf.gif) to win Superbowl 2021, each of the candidates in the Georgia Senate races are about equally likely to win, and if you've always been an apple-a-day person, there's a good chance you ate one on April 17, 2009.  So it still makes sense to talk about probability in uncertain, but not necessarily repeated, situations.

However, the *meaning* of probability does seem different in a physically repeatable situations like coin flips than in single occurrences like the 2021 Superbowl.  Let's switch sports and consider the 2020 World Series of Major League Baseball^[which already happened, but I'm too lazy to update the examples.  Imagine that when you're reading this it's Sept 3, 2020.].  As of Sept 3,

- According to [FiveThirtyEight](https://projects.fivethirtyeight.com/2020-mlb-predictions/1/34), the Los Angeles Dodgers have a 31% chance of winning the 2020 World Series, the highest of any team, while the New York Yankees have a 10% chance.
- According to [FanGraphs](https://www.fangraphs.com/standings/playoff-odds/fg/mlb?date=2020-09-03&dateDelta=), the Dodgers have a 18% chance of winning the 2020 World Series, while the Yankees have an 6% chance.
- According to gambling site [Odds Shark](https://www.oddsshark.com/mlb/world-series-odds), the Dodgers have a 22% chance of winning the 2020 World Series, while the Yankees have an 18% chance.


Each website, as well as many others, assigns different probabilities to the Dodgers or Yankees winning. Which website, if any, is "correct"?

When the situation involves a fair coin flip, we could perform a simulation to see that the long run proportion of flips that land on H is 0.5, and so the probability that a fair coin flip lands on H is 0.5.  Even though the actual 2020 World Series will only happen once, we could still perform a simulation involving hypothetical repetitions. However, simulating the World Series involves first simulating the 2020 season to determine the playoff matchups, then simulating the playoffs to see which teams make the World Series, then simulating the World Series matchup itself.  And simulating the 2020 season involves simulating all the individual games. Even just simulating a single game involves many assumptions; differences in opinions with regards to these assumptions can lead to different probabilities.   For example, on Sept 3, according to FiveThirtyEight the Dodgers had a 68% chance of beating the Colorado Rockies in their game on Sept 4, but according to FanGraphs it was 67% and according to Odds Shark it was 72%.  (The Dodgers won.)  Even though these differences might seem small, many small differences over the course of the season could result in large differences in predictions for the World Series champion.

Unlike physically repeatable situations such as flipping a coin, there is no single set of "rules" for conducting a simulation of a season of baseball games or the World Series champion. Therefore, there is no single relative frequency that determines the probability.  Instead we consider *subjective probability*.

A **subjective (a.k.a. personal) probability** describes the degree of likelihood a given individual assigns to a certain event. As the name suggests, different individuals (or probabilistic models) might have different subjective probabilities for the same event.  In contrast, in the long run relative frequency interpretation the probability is agreed to be defined as the long run relative frequency, a single  number.

**Think of subjective probabilities as measuring *relative degrees of likelihood or uncertainty* ** rather than long run relative frequencies.  For example, in the FiveThirtyEight forecast, the Dodgers are *3.1 times more likely* to win the World Series than the Yankees (3.1 = 0.31 / 0.10).  Relative likelihoods can also be compared across different forecasts or scenarios. For example, FiveThirtyEight believes that the Dodgers are about 1.4 times more likely to win the World Series than Odds Shark does.  Also, FiveThirtyEight believes that the likelihood that a fair coin lands on H is about 5 times larger than the likelihood that the Yankees win the 2020 World Series.

The [FiveThirtyEight MLB predictions](https://fivethirtyeight.com/features/how-our-mlb-predictions-work/) are the output of a probabilistic forecast. A **probabilistic forecast** combines observed data and statistical models to make predictions. Rather than providing a single prediction (such as "the Los Angeles Dodgers will win the 2020 World Series"), probabilistic forecasts provide a range of scenarios and their relative likelihoods.  Such forecasts are subjective in nature, relying upon the data used and assumptions of the model. Changing the data or assumptions can result in different forecasts and probabilities.  In particular, probabilistic forecasts are usually revised over time as more data becomes available.


Simulations can also be based on subjective probabilities.  If we were to conduct a simulation consistent with FiveThirtyEight's model (as of Sept 3), then in about 31% of repetitions the Dodgers would win the World Series, and in about 10% of repetitions the Yankees would win.  Of course, different sets of subjective probabilities correspond to different assumptions and different ways of conducting the simulation.  

Subjective probabilities can be calibrated by weighing the relative favorability of different bets, as in the following example.


```{example subjective-bet}

What is your subjective probability that Professor Ross has a TikTok account? Consider the following two bets, and suppse you can choose only one.

A)  You win $100 if Professor Ross has a TikTok account, and you win nothing otherwise.
A)  A box contains 40 green and 60 gold marbles that are otherwise identical.  The marbles are thoroughly mixed and one marble is selected at random. You win $100 if the selected marble is green, and you win nothing otherwise.

1. Which of the above bets would you prefer?  Or are you completely indifferent? What does this say about your subjective probability that Professor Ross has a Tik Tok account?
1. If you preferred bet B to bet A, consider bet C which has a similar setup to B but now there are 20 green and 80 gold marbles. Do you prefer bet A or bet C? What does this say about your subjective probability that Professor Ross has a Tik Tok account? 
1. If you preferred bet A to bet B, consider bet D which has a similar setup to B but now there are 60 green and 40 gold marbles. Do you prefer bet A or bet D? What does this say about your subjective probability that Professor Ross has a Tik Tok account?
1. Continue to consider different numbers of green and gold marbles.  Can you zero in on your subjective probability?
  
```


```{solution subjective-bet-sol}

to Example \@ref(exm:subjective-bet)

```

```{asis, fold.chunk = TRUE}

1. Since the two bets have the same payouts, you should prefer the one that gives you a greater chance of winning! If you choose bet B you have a 40% chance of winning.
    - If you prefer bet B to bet A, then your subjective probability that Professor Ross has a TikTok account is less than 40%.
    - If you prefer bet A to bet B, then your subjective probability that Professor Ross has a TikTok account is greater than 40%.
    - If you're indifferent between bets A and B, then your subjective probability that Professor Ross has a TikTok account is equal to 40%.  
1. If you choose bet C you have a 20% chance of winning.
    - If you prefer bet C to bet A, then your subjective probability that Professor Ross has a TikTok account is less than 20%.
    - If you prefer bet A to bet C, then your subjective probability that Professor Ross has a TikTok account is greater than 20%.
    - If you're indifferent between bets A and C, then your subjective probability that Professor Ross has a TikTok account is equal to 20%.  
1. If you choose bet D you have a 60% chance of winning.
    - If you prefer bet D to bet A, then your subjective probability that Professor Ross has a TikTok account is less than 60%.
    - If you prefer bet A to bet D, then your subjective probability that Professor Ross has a TikTok account is greater than 60%.
    - If you're indifferent between bets A and D, then your subjective probability that Professor Ross has a TikTok account is equal to 60%.  
1. Continuing in this way you can narrow down your subjective probability. For example, if you prefer bet B to bet A and bet A to bet C, your subjective probability is between 20% and 40%.  Then you  might consider bet E corresponding to 30 gold marbles and 70 green to determine if you subjective probability is greater than or less than 30%.  At some point it will be hard to choose, and you will be in the ballpark of your subjective probability.  (Think of it like going to the eye doctor: "which is better: 1 or 2?"  At some point you can't really see a difference.)



```


Of course, the strategy in the above example isn't an exact science, and there is a lot of behavioral psychology behind how people make choices in situations like this.  But the example gives a very rough idea of how you might discern a subjective probability of an event.


Disclaimer: we do not advocate gambling.  We merely use gambling contexts to motivate probability concepts.


## Working with probabilities {#consistency}


In the previous section we saw two different interpretations of probability: long run relative frequency and subjective.  Fortunately, the mathematics of probability work the same way regardless of the interpretation.  Furthermore, even with subjective probabilities it is helpful to consider what might happen in a simulation.

### Consistency requirements

With either the long run relative frequency or subjective probability interpretation there are some basic logical consistency requirements which probabilities need to satisfy. Put loosely, probabilities cannot be negative and the sum of probabilities over all possible outcomes must be 100%.



```{example worldseries}

As of Sept 3, [FiveThirtyEight](https://projects.fivethirtyeight.com/2019mlb-predictions/) listed the following probabilities for who will 
win the 2020 World Series.

```

| Team                | Probability |
|---------------------|------------:|
| Los Angeles Dodgers |         31% |
| New York Yankees    |         10% |
| Houston Astros      |          9% |
| Tampa Bay Rays      |          9% |
| Other               |             |

According to FiveThirtyEight (as of Sept 3):

1. What would you expect the results of 10000 repetitions of a simulation of the World Series champion to look like?  Construct a table summarizing what you expect. Is this necessarily what would happen? 
1. What must be the probability that the Dodgers do *not* win the 2020 World Series?
1. What must be the probability that one of the above four teams is the World Series champion?
1. What must be the probability that a team other than the above four teams is the World Series champion?  That is, what value goes in the "Other" row in the table?

```{solution worldseries-sol}

to Example \@ref(exm:worldseries)

```

```{asis, fold.chunk = TRUE}


1. While these particular probabilities are subjective, imagining probabilities as relative frequencies often helps our intuition.  If we think of this as a simulation, each repetition results in a World Series champion and in the long run we would expect the Dodgers would be the champion in 31%, or 3100, of the 10000 repetitions.  

    | Team                | Winner |
    |---------------------|-------:|
    | Los Angeles Dodgers |   3100 |
    | New York Yankees    |   1000 |
    | Houston Astros      |    900 |
    | Tampa Bay Rays      |    900 |
    | Other               |   4100 |
    | Total               |  10000 |
  
    Of course, there would be some variability from simulation to simulation, just like in the sets of 1000 coin flips in Figure \@ref(fig:coin-flips-plot3).  But the above counts represent about what we would expect.  

1. 69%. Either the Dodgers win or they don't; if there's a 31% chance that the Dodgers win, there must be a 69% chance that they do not win. If we think of this as a simulation with 10000 repetitions, each repetition results in either the Dodgers winning or not, so if they win in 3100 of repetitions then they must not win in the other 6900.
1. 59%. There is only one World Series champion, so if say the Dodgers win then no other team can win.  Thinking again of the simulation, the repetitions in which the Dodgers win are distinct from those in which the Yankees win.  So if the Dodgers win in 3100 repetitions and the Yankees win in 1000 repetitions, then on a total of 4100 repetitions either the Dodgers or Yankees win.  Adding the four probabilities, we see that the probability that one of the four teams above wins must be 59%.
1. 41%. Either one of the four teams above wins, or some other team wins.  If one of the four teams above wins in 5900 repetitions, then in 4100 repetitions the winner is not one of these four teams.

``` 


```{example worldseries-proportional}

Suppose your subjective probabilities for the 2020 World Series champion satisfy the following conditions.

- The Astros and Rays are equally likely to win
- The Yankees are 1.5 times more likely than the Astros to win
- The Dodgers are 2 times more likely than the Yankees to win
- The winner is as likely to be among these four teams --- Dodgers, Yankees, Astros, Rays --- as not


Construct a table like in Example \@ref(exm:worldseries) of your subjective probabilities.

```


```{solution worldseries-proportional-sol}

to Example \@ref(exm:worldseries-proportional)

```

```{asis, fold.chunk = TRUE}

Here, probabilities are specified indirectly via relative likelihoods.  We need to find probabilities that are in the given ratios and add up to 100%.  It helps to designate one outcome as the "baseline".  It doesn't matter which one; we'll choose the Astros.

- Suppose the Astros account for 1 "unit".  It doesn't really matter what a unit is, but let's say it corresponds to 1000 repetitions of the simulation.  That is, the Astros win in 1000 repetitions.  Careful: we haven't yet specified how many total repetitions we have done, or how many units the entire simulation accounts for.  We're just starting with a baseline of what happens for the Astros.
- The Astro and Rays are equally like to win, so the Rays also account for 1 unit.
- The Yankees are 1.5 times more likely than the Astros to win, so the Yankees account for 1.5 units.  If 1 unit is 1000 repetitions, then the Yankees win in 1500 repetitions, 1.5 times more often than the Astros.
- The Dodgers are 2 times more likely than the Yankees to win, so the Dodgers account for $2\times 1.5=3$ units. If 1 unit is 1000 repetitions, then the Dodgers win in 3000 repetitions.
- The four teams account for a total of $1+1+1.5+3 = 6.5$ units. Since the winner is as likely to among these four teams as not, then "Other" also accounts for 6.5 units.
- In total, there are 13 units which account for 100% of the probability.  The Astros account for 1 unit, so their probability of winning is $1/13$ or about 7.7%.  Likewise, the probability that the Dodgers win is $3/13$ or about 23.1%.


| Team                | Units | Repetitions | Probability |
|---------------------|------:|------------:|------------:|
| Los Angeles Dodgers |   3.0 |        3000 |       23.1% |
| New York Yankees    |   1.5 |        1500 |       11.5% |
| Houston Astros      |   1.0 |        1000 |        7.7% |
| Tampa Bay Rays      |   1.0 |        1000 |        7.7% |
| Other               |   6.5 |        6500 |       50.0% |
| Total               |  13.0 |       13000 |      100.0% |
  
You should verify that all of the probabilities are in the specified ratios.  For example, the Dodgers are 2 times more likely ($2 = 23.1 / 11.5$) than the Yankees to win, and the Yankees are 1.5 times more likely ($1.5 \approx 11.5 / 7.7$) than the Astros to win.

We could have also solved this problem using algebra.  Let $x$ be the probability, as a decimal, that the Astros are the winner.  Then $x$ is also the probability that the Rays are the winner, $1.5x$ for the Yankees, and $3x$ for the Dodgers.  The probability that one of the four teams wins is $x + x + 1.5x + 3x = 6.5x$, so the probability of Other is also  $6.5x$.  The probabilities in decimal form must sum to 1 (that is, 100%), so $1 = x + x + 1.5x + 3x + 6.5x = 13x$.  Solve for $x=1/13$ and then plug in $x=1/13$ to find the other probabilities.


``` 

Example \@ref(exm:worldseries-proportional) illustrates one way of formulating subjective probabilities.  We start by specifying probabilities in relative terms, and then "normalize" these probabilities so that they add up to 100%.  As in the problem, it helps to consider one outcome as a "baseline" and to specify all likelihoods relative to the baseline. 

### Odds

The words "probability", "chance", "likelihood", and "odds" are colloquially treated as synonyms.  However, in the mathematical language of probability, *odds* provide a different way of reporting a probability.  Rather than reporting probability on a 0\% to 100\% scale, odds report probabilities in terms of ratios.  


```{example worldseries-odds}

In Example \@ref(exm:worldseries) the odds that the Yankees win the World Series are 9 to 1 against.

```


1. What do you think that "9 to 1 against" means?
1. What are the odds of the Yankees *not* winning?
1. What are the odds of the Dodgers winning?
1. What are the odds of one of the Other teams winning?
1. The Philadelphia Phillies have 50 to 1 odds against winning.  What is the probability that the Phillies win the World Series?


```{solution worldseries-odds-sol}

to Example \@ref(exm:worldseries-odds)

```



```{asis, fold.chunk = TRUE}

1. The probability that the Yankees win is 0.1, so the probability that they do not win is 0.9.  These numbers are in a 9 to 1 ratio: the probability of not winning (0.9) is 9 times greater than the probability of winning (0.1).  So the odds *against* the Yankees winning the World Series are 9 to 1; "against" because the Yankees are less likely to win than to not win.
1. The probabilities are still in the 9 to 1 ratio, but we can say that the odds are 9 to 1 *in favor* of the Yankees *not* winning.  We could also say the odds are 1 to 9 in favor of the Yankees winning, but odds are typically reported with the larger value first --- 9 to 1 instead of 1 to 9.
1. The probability that the Dodgers win is 0.31 and that they don't win is 0.69, and $0.69/0.31 \approx 2.2$.  So the odds are 2.2 to 1 *against* the Yankees winning; "against" because the Yankees are less likely to win than to not win.  Odds are often reported as whole numbers, so we could say the odds are 11 to 5 against the Dodgers winning.  
1. The probability that an Other team wins is 0.41 and that an Other team doesn't win is 0.59, and $0.59/0.41 \approx 1.44$.  So the odds are 1.44 to 1 against an Other team winning.
1. The probability that the Phillies do not win is 50 times greater than the probability that they do win.  Let the event that the Phillies win account for 1 "unit" so that the event that they do not win accounts for 50 units, for a total of 51 units.  So the probability that the Phillies win is $1/51\approx 0.02$. Note that the probability of not winning, 50/51, is 50 times greater than the probability of winning.

    You could also solve this with algebra.  Let $x$ be the probability that the Phillies win, so $50x$ is the probability that they don't win. The probabilities must sum to 1, so set $x + 50x = 1$ and solve for $x$.


```



The **odds** of an event is a ratio involving the probability that the
event occurs and the probability that the event does not occur.  Odds can be expressed as either "in favor" of or "against" the event occurring.

$$
\begin{aligned}
\text{odds in favor} & = \frac{\text{probability that the event occurs}}{\text{probability that the event does not occur}} \\
& \\
\text{odds against} & = \frac{\text{probability that the event does not occur}}{\text{probability that the event occurs}}\end{aligned}
$$

In some situations odds are typically reported as odds against.  While the odds of an event is a just a single number, odds are often reported as a ratio of whole numbers, e.g., 11 to 1, 7 to 2.  In Example \@ref(exm:worldseries-odds) the odds against the Dodgers winning the World Series are 2.2, which can be reported as 11 to 5 odds (against).

As discussed at the end of Section \@ref(subjective-probability) bets can be used to discern probabilities or odds.

```{example bet-EV}

Ron and Leslie agree to the following bet. They'll ask Professor Ross if he has a TikTok account. If he does, Leslie will pay
Ron $200; if not, Ron will pay Leslie $100.  (Neither has any prior information about whether or not Professor Ross has a TikTok account.)

1. Given this setup, which of the following is being judged as more likely: that Professor Ross has a TikTok account, or that he does not?  Why?
1. What are this bet's odds?
1. Ron and Leslie agree that this is a fair bet, and neither would accept worse odds. What is the subjective probability that Professor Ross has a TikTok account?
1. Suppose they were to hypothetically repeat this bet many times, say 3000 times.  Given the probability from the previous part, how many times would you expect Leslie to win? To lose? What would you expect Leslie's net dollar winnings to be?  In what sense is this bet "fair"?  (Remember: Leslie's winnings are Ron's losses and vice versa.)

```


<!-- 1. Leslie would not agree to the bet if her potential payout were less than $100 given her potential loss of $200. What can you say about Leslie's subjective probability that Professor Ross has a TikTok account? -->
<!-- 1. Ron would not agree to the bet if his potential payout were less than $200 given his potential loss of $100. What can you say about Ron's subjective probability that Professor Ross has a TikTok account? -->


```{solution bet-EV-sol}

to Example \@ref(exm:bet-EV)

```



```{asis, fold.chunk = TRUE}

1. The larger potential payout corresponds to the *less* likely event.  So Professor Ross is more likely to *not* have a TikTok account than to have one.
1. The payouts are in a 2 to 1 ratio, so the odds that Professor Ross has a TikTok account are 2 to 1 against.
1. The odds that Professor Ross has a TikTok account are 2 to 1 against, so Professor Ross is twice as likely to not have a TikTok account than to have one.  This corresponds to a subjective probability^[Technically, Ron and Leslie could still have different subjective probabilities.  Leslie would not agree to worse odds, but she would accept better if Ron offered them.  For example, given a potential loss of $200, Leslie would also agree to a potential payout from Ron of $125 rather than $100. That is, Leslie would accept odds of 1.6 to 1 against ($200/125 = 1.6$), corresponding to a subjective probability of 0.385 ($1/(1 + 1.6)$) .  So Leslie's subjective probability that Professor Ross has a TikTok account is *at least* 1/3.  Similarly, Ron's subjective probability that Professor Ross has a TikTok account is *at most* 1/3.] that Professor Ross has a TikTok account of 1/3 (and a probability that he does not have one of 2/3).
1. The probability that Leslie wins is 1/3, so you would expect her to win in 1000 of the 3000 repetitions.  She wins $200 each time she wins, so you would expect her to win a total of $200,000 on games she wins.  The probability that she loses is 2/3, so you would expect her to lose in 2000 of the 3000 repetions. She loses $100 each time, so you would expect her to lose a total of $200,000 on the games she wins.  So you would expect Leslie's net winnings to be 0, and likewise for Ron.  The bet is fair in the sense that neither party is expected to profit or lose in the long run.


```


The previous example illustrates that the odds of a fair bet on whether or not an event will occur imply a
probability for the event.

\begin{align*}
\text{probability that event occurs} & = \frac{\text{odds in favor of the event}}{1+\text{odds in favor of the event}}\\
& \\
& = \frac{1}{1+\text{odds against the event}}
\end{align*}





<!-- ### Dutch book -->

Regardless of the interpretation --- long run relative frequency or subjective --- probabilities must follow basic logical consistency requirements. If these requirements are mistakenly not satisfied, bad things can happen.  



```{example dutch}

Donny Don't thinks the Dodgers have a pretty good chance to win the World Series.  He thinks their only real competition is the Yankees.  The following are Donny's subjective probabilities for which team will win the World Series.

| Team                | Probability |
|---------------------|------------:|
| Los Angeles Dodgers |         50% |
| New York Yankees    |         25% |
| Other               |         10% |

```

1. What is wrong with Donny's probabilities?
1. What are Donny's odds that the Dodgers win? (Consider only Donny's probability that the Dodgers win.)
1. Would Donny agree to a bet where he pays you $100 if the Dodgers win but you pay him $100 if the Dodgers do not win?
1. What are Donny's odds that the Yankees win? Would Donny agree to a bet where he pays you $150 if the Yankees win but you pay him $50 if the Yankees do not win?
1. What are Donny's odds that a team other than the Dodgers or Yankees wins? Would Donny agree to a bet where he pays you $180 if an other team wins but you pay him $20 if the winner is either the Yankees or Dodgers?
1. Suppose you and Donny agree to make all of the bets in the three previous parts.  Consider your net profit for each of the outcomes (Dodgers win, Yankees win, other wins).  What do you notice?





```{solution dutch-sol}

to Example \@ref(exm:dutch)

```



```{asis, fold.chunk = TRUE}

1. Donny's probabilities do not add up to 100%.
1. Donny's odds that the Dodgers win are $\frac{0.5}{0.5}=1$, or even odds
1. Donny believes that the Dodgers are equally likely to win as to not win so, yes, he would agree to this bet.
1. Donny's odds that the Yankees do not win are $\frac{0.75}{0.25}=3$, or 3 to 1 odds against the Yankees winning. Donny believes that the Yankees are 3 times more likely to not win than to win.  Since the payouts are in a 3 to 1 ratio with the larger payout corresponding to the Yankees winning (the less likely event), then Donny would agree to this bet.
1. Donny's odds that an other team does not win are $\frac{0.9}{0.1}=9$, or 9 to 1 odds against an other team winning. Donny believes that an other team is 9 times more likely to not win than to win.  Since the payouts are in a 9 to 1 ratio with the larger payout corresponding to an other team winning (the less likely event), then Donny would agree to this bet.
1. Given Donny's odds for each outcome, he would agree to each of these bets.

    - If the Dodgers win, you win the first bet but lose the other two, so your net profit is 100 - 50 - 20 = 30.
    - If the Yankees win, you win the second bet but lose the other two, so your net profit is 150 - 100 - 20 = 30
    - If an other team wins, you win the third bet but lose the other two, so your net profit is 180 - 100 - 50 = 30.
    
    Regardless of the outcome, you are guaranteed to earn a net profit of $30.  That's free money for you with no risk, and pretty bad business on Donny's part.  



```

The previous problem contained an example of a "Dutch book". A **Dutch book**^["Book" in the sense of a bookie taking bets.] is a set of probabilities and bets which guarantees a profit, regardless of the outcome of the gamble.  Probabilities that fail to satisfy logical consistency requirements allow for the possibility of Dutch books. The fact that no one should ever want to get caught in a Dutch book, like Donny was in the previous problem, is used to justify why even subjective probabilities should satisfy logical consistency requirements.


## Interpretations of Statistics {#interpretations-of-statistics}

In the previous sections we have seen two interpretations of statistics: relative frequency and subjective. These two interpretations provide the philosophical foundation for two schools of statistics: *frequentist* (hypothesis tests and confidence intervals that you've seen before) and *Bayesian*.  This section provides a very brief introduction to some of the main ideas in Bayesian statistics.  The examples in this section only motivate ideas.  We will fill in lots more details throughout the course.



```{example instructor-age}

How old do you think your instructor (Professor Ross) currently is^[You could probably get a pretty good idea by searching online, but don't do that.  Instead, answer the questions based on what you already know about me.]? Consider age on a
continuous scale, e.g., you might be 20.73 or 21.36 or 19.50.

In this example, you will use probability to quantify your uncertainty about your instructor's age.  You only need to give ballpark estimates of your subjective probabilities, but you might consider what kinds of bets you would be willing to accept like in Example \@ref(exm:subjective-bet).  (This exercise just motivates some ideas.  We'll fill in lots of details later.)

```



1.  What is your subjective probability that your instructor is at most 30 years old? More than 30 years old? (What must be true about these two probabilities?)
1.  What is your subjective probability that your instructor is at most 60 years old? More than 60 years old?
1.  What is your subjective probability that your instructor is at most 40 years old? More than 40 years old?
1.  What is your subjective probability that your instructor is at most 50 years old? More than 50 years old?
1. Fill in the blank: your subjective probability that your instructor is at most [blank] years old is equal to 50%.
1. Fill in the blanks: your subjective probability that your instructor is between [blank] and [blank] years old is equal to 95%.
1.  Let $\theta$ represent your instructor's age at midnight on Jan 4, 2021. Use your answers to the previous parts to sketch a
    continuous probability density function to represent your
    *subjective probability distribution* for $\theta$.
1.  If you ascribe a probability distribution to $\theta$, then are you treating $\theta$ as a constant or a random variable?
    



```{solution intructor-age-sol}

to Example \@ref(exm:instructor-age)

```



```{asis, fold.chunk = TRUE}

Even though in  reality your instructor's current age is a fixed number, its value is unknown or uncertain to you, and you can use probability to quantify this uncertainty.  You would probably be willing to bet any amount of money that your instructor is over 20 years old, so you would assign a probability of 100% to that event, and 0% to the event that he's at most 20 years old.  Let's say you're pretty sure that he's over 30, but you don't know that for a fact, so you assign a probability of 99% to that event (and 1% to the event that he's at most 30).  You think he's over 40, but you're even less sure about that, so maybe you assign the event that he's over 40 a probability of 67% (say you'd accept a bet at 2 to 1 odds.)  You think there's a 50/50 chance that he's over 50.  You're 95% sure that he's between 35 and 60.  And so on.  Continuing in this way, you can start to determine a probability distribution to represent your beliefs about the instructor's age.  Your distribution should correspond to your subjective probabilities.  For example, the distribution should assign a probability of 67% to values over 40.

This is just one example.  Different students will have different distributions depending upon (1) how much information you know about the instructor, and (2) how that information informs your beliefs about the instructor's age.  We'll see some example plots in the next exercise.

Regarding the last question, since we are using a probability distribution to quantify our uncertainty about $\theta$, we are treating $\theta$ as a *random variable*.  

```



Recall that a **random variable** is a numerical quantity whose value is determined by the outcome of a random or uncertain phenomenon.
The random phenomenon might involve physically repeatable randomness, as in "flip a coin 10 times and count the number of heads." But remember that "random" just means "uncertain" and there are lots of different kinds of uncertainty.  For example, the total number of points scored in the 2021 Superbowl will be one and only one number, but since we don't know what that number is we can treat it as a random variable.  Treating the number of points as a random variable allows us to quantify our uncertainty about it through probability statements like "there is a 50% chance that fewer than 45 points will be scored in Superbowl 2021".

The **(probability) distribution** of a random variable specifies the possible values of the random variable and a way of determining corresponding probabilities. Like probabilities themselves, probability distributions of random variables can also be interpreted as:

- *relative frequency distributions*, e.g., what pattern would emerge if I simulated many values of the random variable? or as
- *subjective probability distributions*, e.g., which potential values of this uncertain quantity are relatively more plausible than others?

As the name suggests, different individuals might have different subjective probability distributions for the same random variable.

```{example instructor-age2}

Continuing Example \@ref(exm:instructor-age), the plot below displays the subjective probability distribution of the instructor's age of four students.  

```




```{python, echo = FALSE}

Uniform(25, 75).plot()
Normal(45, 3).plot()
(25 + 50 * RV(Beta(3, 6))).sim(10000).plot('density')
(25 + 50 * RV(Beta(3, 3))).sim(10000).plot('density')

plt.xlabel('Instructor age');
plt.ylabel('Density');
plt.legend(['Ariana', 'Billie', 'Cardi', 'Dua']);
plt.show();

```

1. Since age is treated as a continuous random variable, each of the above plots is a probability "density".  Explain briefly what this means.  How is probability represented in density plots like these?
1. Rank the students in terms of their subjective probability that the instructor is at most 40 years old.
1. Rank the students in terms of their answers to the question: your subjective probability that your instructor is at most [blank] years old is equal to 50%.
1. Rank the students in terms of their uncertainty about the instructor's age.  Who is the most uncertain?  The least?

```{solution intructor-age2-sol}

to Example \@ref(exm:instructor-age2)

```



```{asis, fold.chunk = TRUE}

1. In a density plot, probability is represented by area under the curve.  The total area under each curve is 1, corresponding to 100% probability.  The density height at any particular value $x$ represents the relative likelihood that the random variable takes a value "close to" $x$.  (We'll consider densities in more detail later.)
1. Each student's subjective probability that the instructor is at most 40 is equal to the area under her subjective probability density over the range of values less than 40. Billie has the smallest probability, then Dua, then Ariana, then Cardi has the largest probability.
1. Now we want to find the "equal areas point" of each distribution.  From smallest to largest: Cardi then Billie, and Ariana and Dua appear to be about the same.  The equal areas point appears to be around 40 or so for Cardi.  It's definitely less than 45, which appears to the equal areas point for Billie.  The equal areas point for Ariana is 50 (halfway between 25 and 75), and Dua's appears to be about 50 also.
1. Ariana is most uncertain, then Dua, then Cardi, then Billie is the least uncertain. Each distribution represents 100% probability, but Ariana stretches this probability over the largest range of possibe values, while Billie stretches this over the shortest.  Ariana is basically saying the instructor can be any age between 25 and 75.  Billie is fairly certain that the instructor is close to 45, and she's basically 100% certain that the instructor is between 35 and 55.

```

    



The previous examples introduce how probability can be used to quantify uncertainty about unknown numbers.  One key aspect of Bayesian analyses is applying a subjective probability distribution to a *parameter* in a statistical model.


```{example, harry-potter}

Let $\theta_b$ represent the proportion of current Cal Poly students who have ever read any of the books in the *Harry Potter* series.  Let $\theta_m$ represent the proportion of current Cal Poly students who have ever seen any of the movies in the *Harry Potter* series.

```


1. Are $\theta_b$ and $\theta_m$ parameters or statistics?  Why?
1. Are the values of $\theta_b$ and $\theta_m$ known or unknown, certain or uncertain?
1. What are the possible values of $\theta_b$ and $\theta_m$?
1. Sketch a probability distribution representing what you think are more/less credible values of $\theta_b$.  Repeat for $\theta_m$.  Are you more certain about the value of $\theta_b$ or $\theta_m$; how is this reflected in your distributions?
1. Suppose that in a class of 35 Cal Poly students, 21 have read a Harry Potter book, and 30 have seen a Harry Potter movie.  Now that we have observed some data, sketch a probability distribution representing what you think are more/less credible values of $\theta_b$.  Repeat for $\theta_m$. How do your distributions after observing data compare to the distributions you sketched before? 



```{solution harry-potter-sol}

to Example \@ref(exm:harry-potter)

```



```{asis, fold.chunk = TRUE}

1. The population of interest is current Cal Poly students, so $\theta_b$ and $\theta_m$ are *parameters*.  We don't have relevant data for the entire population, but we could collect data on a sample.
1. Since we don't have data on the entire population, the values of $\theta_b$ and $\theta_m$ are unknown, uncertain.
1. $\theta_b$ and $\theta_m$ are proportions so they take values between 0 and 1.  Any value on the continuous scale between 0 and 1 is theoretically possible, though the values are not equally plausible.
1. Results will vary, but here's my thought process.  I think that a strong majority of Cal Poly students have seen at least one Harry Potter movie, maybe 80% or so. I wouldn't be that surprised if it were even close to 100%, but I would be pretty surprised if it were less than 60%.  

    However, I'm less certain about $\theta_b$.  I suspect that fewer than 50% of students have read at least one Harry Potter book, but I'm not very sure and I wouldn't be too surprised if it were actually more than 50%.  
    
    See the figure on the left in \@ref(fig:plot-harry-potter) for what my subjective probability distributions might look like.  Since I am more uncertain about $\theta_b$, its density is "spread out" over a wider range of values.
1. The values of $\theta_b$ and $\theta_m$ are still unknown, but I am less uncertain about their values now that I have observed some data.  The sample proportion who have watched a Harry Potter movie is $30/35 = 0.857$, which is pretty consistent with my initial beliefs.  But now I update my subjective distribution to concentrate even more of my subjective probability on values in the 80 percent range.  

    I had suspected that $\theta_b$ was less than 0.5, so the observed sample proportion of $21/35 = 0.6$ goes against my expectations.  However, I was fairly uncertain about the value of $\theta_m$ prior to observing the data, so 0.6 is not too surprising to me.  I update my subjective distribution so that it's centered closer to 0.6, while still allowing for my suspicion that $\theta_b$ is less than 0.5.  
    
    See the figure on the right in \@ref(fig:plot-harry-potter) for what my subjective probability distributions might look like after observing the sample data. Of course, the sample proportions are not necessarily equal to the population proportions.  But if the samples are reasonably representative, I would hope that the observed sample proportions are close to the respective population proportions.  Even after observing data, there is still uncertainty about the parameters $\theta_b$ and $\theta_m$, and my subjective distributions quantify this uncertainty.
  
```

(ref:cap-harry-potter) Example subjective distributions in Example \@ref(exm:harry-potter). Left: prior to observing sample data.  Right: after observing sample data.

```{r, plot-harry-potter, fig.cap='(ref:cap-harry-potter)', echo=FALSE, fig.show="hold", out.width="50%"}

theta <- seq(0, 1, length = 10000)

# prior

a_m = 15
b_m = 4
pi_m = dbeta(theta, a_m, b_m)

a_b = 3
b_b = 3.6
pi_b = dbeta(theta, a_b, b_b)

plot(theta, pi_m, type = "l", lty = 1, lwd = 2, col = "orange",
     xlab = "Population proportion",
     ylab = "Density",
     yaxt = "n",
     main = "Prior distributions")

lines(theta, pi_b, lwd = 2, col = "skyblue")

legend("topleft", c("movie", "book"), lty = c(1, 1), lwd = c(2, 2),
       col = c("orange", "skyblue"))

# posterior
n = 35
y_b = 21
y_m = 30

a_b1 = a_b + y_b
b_b1 = b_b + n - y_b
pi_b1 = dbeta(theta, a_b1, b_b1)

a_m1 = a_m + y_m
b_m1 = b_m + n - y_m
pi_m1 = dbeta(theta, a_m1, b_m1)

plot(theta, pi_m1, type = "l", lty = 1, lwd = 2, col = "orange",
     xlab = "Population proportion",
     ylab = "Density",
     yaxt = "n",
     main = "Posterior distributions")

lines(theta, pi_b1, lwd = 2, col = "skyblue")
lines(theta, pi_m, lwd = 1, lty = 2, col = "orange")
lines(theta, pi_b, lwd = 1, lty = 2, col = "skyblue")

legend("topleft",
       c("movie (posterior)", "book (posterior)", "movie (prior)", "book (prior)"),
       lty = c(1, 1, 2, 2), lwd = c(2, 2, 1, 1),
       col = rep(c("orange", "skyblue"), 2))

abline(v = c(y_m / n, y_b / n), lty = 3, col = c("orange", "skyblue"))

```


Recall some statistical terminology.

- **Observational units** (a.k.a., cases, individuals, subjects) are the people, places, things, etc we collect information on.
- A **variable** is any characteristic of an observational unit that we can measure.
- **Statistical inference** involves using data collected on a *sample* to make conclusions about a *population*.  
- Inference often concerns specific numerical summaries, using values of *statistics* to make conclusions about *parameters*.
- A **parameter** is a number that describes the **population**, e.g., *population mean*, *population proportion*.  The actual value of a parameter is almost always *unknown*.
    - Parameters are often denoted with Greek letters.  We'll often use the Greek letter $\theta$ ("theta") to denote a generic parameter.
- A **statistic** is a number that describes  the **sample*, e.g., *sample mean*, *sample proportion*.



Parameters are unknown numbers.  In "traditional", *frequentist* statistical analysis, parameters are treated as *fixed --- that is, not random --- constants*.  Any randomness in a frequentist analysis arises from how the data were collected, e.g., via random sampling or random assignment. In a frequentist analysis, statistics are random variables; parameters are fixed numbers.

For example, a frequentist 95% confidence interval for $\theta_b$ in the previous example is [0.434, 0.766].  We estimate with 95% confidence that the proportion of Cal Poly students that have read any of the books in the Harry Potter series is between 0.434 and 0.766.  Does this mean that there is a 95% probability that $\theta_b$ is between 0.434 and 0.766?  No!  In a frequentist analysis, the parameter $\theta_b$ is treated like a fixed constant.  That constant is either between 0.434 and 0.766 or it's not; we don't know which it is, but there's no probability to it.  In a frequentist analysis, it doesn't make sense to say "what is the probability that $\theta_b$ (a number) is between 0.434 and 0.766?" just like it doesn't make sense to say "what is the probability that 0.5 is between 0.434 and 0.766?"  Remember that 95% confidence derives from the fact that for 95% *of samples* the procedure that was used to produce the interval [0.434, 0.766] will produce intervals that contain the true parameter $\theta_b$.  It is the samples and the intervals that are changing from sample to sample; $\theta_b$ stays constant at its fixed but unknown value.  In a frequentist analysis, probability quantifies the *randomness in the sampling procedure*.

On the other hand, in a Bayesian statistical analysis, since a parameter $\theta$ is unknown --- that is, it's value is *uncertain* to the observer --- $\theta$ is treated as a *random variable*.  That is, **in Bayesian statistical analyses unknown parameters are random variables that have probability distributions.**  The probability distribution of a parameter quantifies the degree of uncertainty about the value of the parameter.  Therefore, the Bayesian perspective allows for probability statements about parameters. For example, a Bayesian analysis of the previous example might conclude that there is a 95% chance that $\theta_b$ is between 0.426 and 0.721.  Such a statement is valid in the Bayesian context, but nonsensical in the frequentist context.

In the previous example, we started with distributions that represented our uncertainty about $\theta_b$ and $\theta_m$ based on our "beliefs", then we revised these distributions after observing some data.  If we were to observe more data, we could revise again.  In this course we will see (among other things) (1) how to quantify uncertainty about parameters using probability distributions, and (2) how to update those distributions to reflect new data.


Throughout these notes we will focus on Bayesian statistical analyses.  We will occasionally compare Bayesian and frequentist analyses and viewpoints. But we want to make clear from the start: Bayesian versus frequentist is NOT a question of right versus wrong.  Both Bayesian and frequentist are valid approaches to statistical analyses, each with advantages and disadvantages.  We'll address some of the issues along the way.  But at no point in your career do you need to make a definitive decision to be a Bayesian or a frequentist; a good modern statistician is probably a [bit of both](https://tenor.com/view/star-lord-bit-of-both-chris-pratt-guardians-of-the-galaxy-peter-quill-gif-11821953).
