# Introduction to Multi-Parameter Models {#multi-parameter}


So far we have considered situations with a just a single unknown parameter $\theta$.  However, many interesting problems involve multiple unknown parameters.

For example, in Chapter \@ref(mean) we considered the problem of estimating the population mean of a numerical variable assuming the population standard deviation was known.
However, in practice both the population mean and population standard deviation are unknown.
Even if we are only interested in estimating the population mean, we still need to account for the uncertainty in the population standard deviation.

When there are two (or more) unknown parameters the prior and posterior distribution will each be a *joint* probability distribution over *pairs* (or tuples/vectors) of possible values of the parameters.




```{example, body-temp-discrete-sd}

Assume body temperatures (degrees Fahrenheit) of healthy adults follow a Normal distribution with unknown mean $\mu$ and unknown standard deviation $\sigma$.
Suppose we wish to estimate both $\mu$, the population mean healthy human body temperature, and $\sigma$, the population standard deviation of body temperatures.  

```

1. Assume a discrete prior distribution according to which

    - $\mu$ takes values 97.6, 98.1, 98.6 with prior probability 0.2, 0.3, 0.5, respectively.
    - $\sigma$ takes values 0.5, 1 with prior probability 0.25, 0.75, respectively.
    - $\mu$ and $\sigma$ are independent.
    
    Start to construct the Bayes table.  What are the possible values of the parameter?  What are the prior probabilities?  (Hint: the parameter $\theta$ is a *pair* $(\mu, \sigma)$.)
    
1. Suppose two temperatures of 97.5 and 97.9 are observed, independently.
Identify the likelihood.
1. Complete the Bayes table and find the posterior distribution after observing these two measurements.  Compare to the prior distribution.
1. Suppose that we only observe that in a sample of size 2 the mean is 97.7.  Is this information enough to evaluate the likelihood function and determine the posterior distribution?
1. The prior assumes that $\mu$ and $\sigma$ are independent.  Are they independent according to the posterior distribution?


```{solution, body-temp-discrete-sd-sol}
to Example \@ref(exm:body-temp-discrete-sd)
```


1. See the table below.
There are 3 possible values for $\mu$ and 2 possible values for $\sigma$ so there are $(3)(2)=6$ possible $(\mu, \sigma)$ pairs.
Each row in the Bayes table represents a $(\mu, \sigma)$ pair.
Since  the prior assumes independence, the prior probability of any pair is the product of the marginal prior probabilities of $\mu$ and $\sigma$.
For example, the probability probability that $\mu = 97.6$ and $\sigma=0.5$ is $(0.2)(0.25)=0.05$

1. The likelihood is similar to what we have seen in other examples concerning body temperature, but it is now a function of both $\mu$ and $\sigma$.
The likelihood is determined by evaluating, for each $(\mu, \sigma)$ pair, the Normal($\mu$, $\sigma$) density at each of $y=97.9$ and $y=97.5$ and then finding the product:
\[
{\scriptstyle
f(y=(97.9, 97.5)|\mu, \sigma)  \propto \left[\sigma^{-2}\,\exp\left(-\frac{1}{2}\left(\frac{97.9-\mu}{\sigma}\right)^2\right)\right]\left[\sigma^{-2}\,\exp\left(-\frac{1}{2}\left(\frac{97.5-\mu}{\sigma}\right)^2\right)\right]
}
\]

1. See the table below. As always, posterior is proportional to likelihood times prior.
For the sample (97.9, 97.5), the sample mean is 97.7 and the sample standard deviation is 0.283.
The posterior distribution pushes probability away from $\mu=98.6$, and pushes more probability towards $\sigma=0.5$.



    ```{r}
# prior
mu = c(97.6, 98.1, 98.6)
sigma = c(0.5, 1)
theta = expand.grid(mu, sigma)
names(theta) = c("mu", "sigma")

prior_mu = c(0.20, 0.30, 0.50)
prior_sigma = c(0.25, 0.75)
prior = apply(expand.grid(prior_mu, prior_sigma), 1, prod)
prior = prior / sum(prior)

# data
y = c(97.9, 97.5) # single observed value

# likelihood
likelihood = dnorm(97.9, mean = theta$mu, sd = theta$sigma) * dnorm(97.5, mean = theta$mu, sd = theta$sigma)

# posterior
product = likelihood * prior
posterior = product / sum(product)

# bayes table
bayes_table = data.frame(theta,
                         prior,
                         likelihood,
                         product,
                         posterior)

kable(bayes_table, digits = 4, align = 'r')

    ```

1. Knowing only the posterior mean would not be sufficient, since it would not give us enough information to estimate the standard deviation $\sigma$.

1. The posterior distribution represents some dependence between $\mu$ and $\sigma$.
For example, consider the pair $\mu=97.6$ and $\sigma=0.5$.
The marginal posterior probability that $\mu=97.6$ is `r round(sum(posterior[c(1, 4)]), 4)`.
The marginal posterior probability that $\sigma=0.5$ is `r round(sum(posterior[1:3]), 4)`.
But the joint posterior probability that $\mu=97.6$ and $\sigma=0.5$ is `r round(posterior[1], 4)`, which is not the product of the marginal probabilities. 







```{example, body-temp-prior-sd}
Continuing  Example \@ref(exm:body-temp-discrete-sd), let's assume a more reasonable, continuous prior for $(\mu,\sigma)$.
We have seen that we often work with the precision $\tau = 1/\sigma^2$ rather than the SD.
Assume a continuous prior distribution which assumes

- $\mu$ has a Normal distribution with mean 98.6 and standard deviation 0.7.
- $\tau$ has a Gamma distribution with shape parameter 5 and rate parameter 2.
- $\mu$ and $\tau$ are independent.

(This problem just concerns the prior distribution. We'll look at this posterior distribution in the next example.)

```


1. Simulate $(\mu, \tau)$ pairs from the prior distribution and plot them.
1. Simulate $(\mu, \sigma)$ pairs from the prior distribution and plot them.
Describe the prior distribution of $\sigma$.
1. Find and interpret a central 95% prior credible interval for $\mu$.
1. Find a central 95% prior credible interval for the precision $\tau=1/\sigma^2$.
1. Find and interpret a central 95% prior credible interval for $\sigma$.
1. What is the prior credibility that both $\mu$ and $\sigma$ lie within their credible intervals?



```{solution}
to Example \@ref(exm:body-temp-prior-sd)
```


1. We could plot the prior distribution directly.
However, distributions are usually only approximated via simulation, so we'll just simulate.
The prior distribution is a distribution on $(\mu,\tau)$ pairs.

    ```{r}
Nrep = 10000
    
mu_sim_prior = rnorm(Nrep, 98.6, 0.7)
tau_sim_prior = rgamma(Nrep, 5, 2)
sigma_sim_prior = 1 / sqrt(tau_sim_prior)
sim_prior = data.frame(mu_sim_prior, tau_sim_prior, sigma_sim_prior)

ggplot(sim_prior, aes(mu_sim_prior, tau_sim_prior)) +
  geom_point(color = "skyblue", alpha = 0.4) +
  geom_density_2d(color = "orange", size = 1)

```


1. See plots below.  The prior distribution on $\tau$ induces a prior distribution on $\sigma = 1/\sqrt{\tau}$.


    ```{r}

ggplot(sim_prior, aes(x = sigma_sim_prior)) + 
 geom_histogram(aes(y=..density..), color = "black", fill = "white") +
 geom_density(size = 1, color = "orange")

ggplot(sim_prior, aes(mu_sim_prior, sigma_sim_prior)) +
  geom_point(color = "skyblue", alpha = 0.4) +
  geom_density_2d(color = "orange", size = 1)

```

1. There is a prior probability of 95% that population mean human body temperature is between `r round(quantile(mu_sim_prior, c(0.025)), 2)` and `r round(quantile(mu_sim_prior, c(0.975)), 2)` degrees F.

    ```{r}
quantile(mu_sim_prior, c(0.025, 0.975))
```


1. We can compute a credible interval like usual.  Precision just doesn't have as practical an interpretation as standard deviation.

    ```{r}
quantile(tau_sim_prior, c(0.025, 0.975))
```

1. There is a prior probability of 95% that population standard deviation of human body temperatures is between `r round(quantile(sigma_sim_prior, c(0.025)), 2)` and `r round(quantile(sigma_sim_prior, c(0.975)), 2)` degrees F.

    ```{r}
quantile(sigma_sim_prior, c(0.025, 0.975))
```

1. Since $\mu$ and $\sigma$ are independent according to the prior distribution, the probability that both parameters lie in their respective intervals is (0.95)(0.95)=0.9025.
If we want 95% joint prior credibility, we need a different region.


```{example, body-temp-post-small}
Continuing the previous example, we'll now compute the posterior distribution given a sample of two measurements of 97.9 and 97.5.

```



1. Assume a grid of $\mu$ values from 96.0 to 101.0 in increments of 0.001, and a grid of $\tau$ values from 0.1 to 8.0 in increments of 0.001.
How many possible values of the pair $(\mu, \tau)$ are there; that is, how many rows are there in the Bayes table?
1. Use grid approximation to approximate the joint posterior distribution of $(\mu, \tau)$
Simulate values from the joint posterior distribution and plot them.
Compute the posterior correlation between $\mu$ and $\tau$; are they independent according to the posterior distribution?
1. Plot the simulated joint posterior distribution of $\mu$ and $\sigma$.  Compare to the prior.
1. Suppose we wanted to approximate the posterior distribution without first using grid approximation.
Describe how, in principle, you would use a naive (non-MCMC) simulation to approximate the posterior distribution.
In practice, what is the problem with such a simulation?


```{solution}
to Example \@ref(exm:body-temp-post-small)
```


1. There are (101-96)/0.001 = 5000 values of $\mu$ in the grid (actually 5001 including both endpoints) and (8-0.1)/0.001 = 7900 values of $\mu$ in the grid (actually 7901).
There are over 39 million possible values of the pair $(\mu, \tau)$ in the grid.
1. See below.  Even though $\mu$ and $\tau$ are independent according to the prior distribution, there is a negative posterior correlation.

    ```{r}
# parameters
mu = seq(96.0, 101.0, 0.001)
tau = seq(0.1, 8, 0.001)

theta = expand.grid(mu, tau)
names(theta) = c("mu", "tau")
theta$sigma = 1 / sqrt(theta$tau)

# prior
prior_mu_mean = 98.6
prior_mu_sd = 0.7

prior_precision_shape = 5
prior_precision_rate = 2

prior = dnorm(theta$mu, prior_mu_mean, sd = prior_mu_sd) *
  dgamma(theta$tau, shape = prior_precision_shape,
                         rate = prior_precision_rate)
prior = prior / sum(prior)

# data
y = c(97.9, 97.5)

# likelihood
likelihood = dnorm(97.9, mean = theta$mu, sd = theta$sigma) *
  dnorm(97.5, mean = theta$mu, sd = theta$sigma)

# posterior
product = likelihood * prior
posterior = product / sum(product)

# posterior simulation
sim_posterior = theta[sample(1:nrow(theta), 10000, replace = TRUE, prob = posterior), ]

#plots 

ggplot(sim_posterior, aes(mu, tau)) +
  geom_point(color = "skyblue", alpha = 0.4) +
  geom_density_2d(color = "orange", size = 1)

cor(sim_posterior$mu, sim_posterior$tau)


```


1. See below. We see that the posterior shifts the density towards smaller values of $\mu$ and $\sigma$.

    ```{r}
ggplot(sim_posterior, aes(mu, sigma)) +
  geom_point(color = "skyblue", alpha = 0.4) +
  geom_density_2d(color = "orange", size = 1)

cor(sim_posterior$mu, sim_posterior$sigma)
```


1. Simulate a value of $(\mu, \sigma)$ from their joint prior distribution, by simulating a value of $\mu$ from a Normal(98.6, 0.7) distribution and simulating a value of $\tau$ from a Gamma(5, 2) distribution and setting $\sigma = 1/\sqrt{\tau}$.
Given $\mu$ and $\sigma$ simulate two independent $y$ values from a Normal($\mu$, $\sigma$) distribution.
Repeat many times.
Condition on the observed data by discarding any repetitions for which the $y$ values are not (97.9, 97.5), to some reasonable degree of precision, say rounded to 1 decimal place.
Approximate the posterior distribution using the remaining simulated values of $(\mu, \sigma)$.

    In practice, the probability of seeing a sample with $y$ values of 97.9 and 97.5 is extremely small, so almost all repetitions of the simulation would be discarded and such a simulation would be extremely computationally inefficient.
(For example, the values of $\mu$ and $\sigma$ which maximize the likelihood of (97.9, 97.5) are 97.7 and 0.2, respectively, and even for those values and rounding to 1 decimal place the probability of seeing such a sample is only 0.015.)



The previous problem illustrates that grid approximation can quickly become computationally infeasible when there are multiple parameters (to obtain sufficient precision).
Naively conditioning a simulation on the observed sample is also computationally infeasible, since except in the simplest situations the probability of recreating the observed sample in a simulation is essentially 0.
Therefore, we need more efficient computational methods, and MCMC will do the trick.


```{example, temp-jags}
The `temperature` data file contains 208 measurements of human body temperature (degrees F).
The sample mean is 97.71 degrees F and the sample SD is 0.75 degrees F.
Assuming the same prior distribution as in the previous problem, use JAGS to approximate the joint posterior distribution of $\mu$ and $\sigma$.
Summarize the posterior distribution in context.
```


```{solution}
to Example \@ref(exm:temp-jags)
```

The JAGS code is below.  A few comments on the code

- The data is read as individual values, so the likelihood of each `y[i]` is computed via a for loop.
- We have called the parameters by their names `mu`, `tau`, `sigma`, rather than just a single `theta`.
- We specify a prior distribution on `tau` and then define `sigma <- 1 / sqrt(tau)`.
- We are interested in the posterior distribution of $\mu$ and $\sigma$, so we include both parameters in the `model.names` argument of the `coda.samples` function.
- The output of `coda.samples` is a special object called an `mcmc.list`.
Calling `plot` on this object produces a trace plot and a density plot for each parameter included in `variable.names`.
But it does not automatically produce any joint distribution plots.
- We use `mcmc_scatter` from the `bayesplot` package to create a scatter plot of the joint posterior distribution, to which we can add contours.
- The simulated values from an `mcmc.list` can be extracted as a matrix with `as.matrix` and then manipulated as usual, e.g., to compute the correlation.

```{r}
    
Nrep = 10000
Nchains = 3

# data
data = read.csv("_data/temperature.csv")
y = data$temperature
n = length(y)

# model
model_string <- "model{

  # Likelihood
  for (i in 1:n){
    y[i] ~ dnorm(mu, 1 / sigma ^ 2)
  }

  # Prior
  mu ~ dnorm(98.6, 1 / 0.7 ^ 2)
  
  sigma <- 1 / sqrt(tau)
  tau ~ dgamma(5, 2)

}"

# Compile the model
dataList = list(y=y, n=n)

model <- jags.model(textConnection(model_string), 
                    data=dataList,
                    n.chains=Nchains)

update(model, 1000, progress.bar="none")

posterior_sample <- coda.samples(model, 
                                 variable.names=c("mu", "sigma"),
                                 n.iter=Nrep,
                                 progress.bar="none")

head(as.matrix(posterior_sample))

# Summarize and check diagnostics
summary(posterior_sample)
plot(posterior_sample)

# Scatterplot from bayesplot package
color_scheme_set("brightblue")
mcmc_scatter(posterior_sample, pars = c("mu", "sigma"), alpha = 0.1) +
  stat_ellipse(level = 0.95, color = "black", size = 2) +
  stat_density_2d(color = "orange", size = 1)

# posterior correlation
cor(as.matrix(posterior_sample))
```

A few comments about the posterior distribution

- The joint posterior distribution appears to be roughly Bivariate Normal.
The correlation is close to 0, indicating independence^[Remember, if $X$ and $Y$ are independent then the correlation is 0, but the converse is not true in general. However, if $X$ and $Y$ have a *Bivariate Normal* distribution and their correlation is 0, then $X$ and $Y$ are independent.] between $\mu$ and $\sigma$ in the posterior.
- The posterior distribution of $\mu$ is approximately Normal with posterior mean 97.71 (basically the sample mean) and posterior SD 0.05.
There is a 95% posterior probability that the population mean human body temperature is between 97.6 and 97.8 degrees F.
- The posterior distribution of $\sigma$ is approximately Normal with posterior mean 0.75 (basically the sample SD) and posterior SD 0.036.
There is a 95% posterior probability that the population SD of human body temperatures is between 0.68 and 0.82 degrees F.
- There is a posterior probability of 90.25% that both of the above statements are true.
To have joint posterior credibility of 95%, we could lengthen each interval (to 97.5% for two independent intervals) to obtain a rectangular credibility region.
The scatterplot also shows a 95% posterior credible ellipse (in black) for both $\mu$ and $\sigma$.


```{example, temp-jags-predict}
Continuing the previous example, how could you use simulation to approximate the posterior predictive distribution of a single body temperature?
Conduct the simulation and compute and interpret a 95% prediction interval.
```

```{solution}
to Example \@ref(exm:temp-jags-predict)
```

- Simulate a $(\mu, \sigma)$ pair from the joint posterior distribution.
- Given $\mu$ and $\sigma$, simulate a value of $y$ from a $N(\mu, \sigma)$ distribution.
- Repeat many times and summarize the simulated $y$ values to approximate the posterior predictive distribution.

See the code below.
JAGS has already returned a simulation from the joint posterior distribution of $(\mu, \sigma)$
For each of these simulated values, simulate a corresponding $y$ value like usual.

```{r}
theta_sim = as.matrix(posterior_sample)

y_sim = rnorm(nrow(theta_sim), theta_sim[, "mu"], theta_sim[, "sigma"])

hist(y_sim, freq = FALSE, xlab = "Body temperature (degrees F)",
     main = "Posterior preditive distribution")
lines(density(y_sim))
abline(v = quantile(y_sim, c(0.025, 0.975)), col = "orange")

quantile(y_sim, c(0.025, 0.975))
```


There is a posterior predictive probability of 95% that a body temperature is between 96.25 and 99.20 degrees F.
Roughly, 95% of healthy human body temperatures are between 96.25 and 99.20 degrees F.
