--- 
title: "An Introduction to Bayesian Reasoning and Methods"
author: "Kevin Ross"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::gitbook:
    split_by: section
    config:
      toc:
        collapse: section
        scroll_highlight: yes
        before: null
        after: null
      toolbar:
        position: fixed
      edit : null
      download: null
      search: yes
      fontsettings:
        theme: white
        family: sans
        size: 2
      sharing:
        facebook: no
        twitter: no
        google: no
        linkedin: no
        weibo: no
        instapper: no
        vk: no
        all: ['facebook', 'google', 'twitter', 'linkedin', 'weibo', 'instapaper']
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bayesian-reasoning-and-methods
description: "This textbook presents an introduction to Bayesian reasoning and methods"
---

# Preface{-}

```{r include=FALSE}

library(knitr)
# library(learnr)
library(bookdown)
library(tidyverse)
library(reticulate)
library(ggplot2)
library(scales)

blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold", hjust = 0.5)
  )

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')

knitr::opts_chunk$set(echo = TRUE)

options(scipen = 999)

# knitr::opts_chunk$set(eval = FALSE)



hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    fold <- options[[paste0("fold.", type)]]
    part = hooks[[type]](x, options)
    
    # If fold is FALSE, don't fold
    if (isFALSE(fold)) return(paste0(part))

    # If fold is TRUE, fold
    else if (isTRUE(fold)) return(paste0(
      "<details><summary>", paste("Show/hide solution", " "), "</summary>\n\n",
      part,
      "\n\n</details>"
    ))
    
    # If fold is not specified, don't fold
    else return (paste0(part))

  }
}
knitr::knit_hooks$set(
  chunk = hook_foldable("chunk")
)


```


```{r, include = FALSE}
py_install("symbulate", pip = TRUE)
```


```{python, include = FALSE}
# These lines needed to show plots because of
# reticulate issue with matplotlib
# https://github.com/rstudio/rstudio/issues/4182

# import os
# os.environ['QT_QPA_PLATFORM_PLUGIN_PATH'] = 'C:/Users/kjross/Anaconda3/Library/plugins/platforms'

from symbulate import *
import matplotlib
import matplotlib.pyplot as plt
# from IPython.display import Markdown, display, HTML

```




(ref:ddwddd) Don't do what Donny Don't does.


<!-- ### Why study Bayesian? {-} -->

<!-- ### Learn by doing {-} -->

This textbook presents an introduction to Bayesian reasoning and methods.

The exercises in this book are used to both motivate new topics and to help you practice your understanding of the material.  You should attempt the exercises on your own before reading the solutions.  To encourage you to do so, the solutions have been hidden.  You can reveal the solution by clicking on the **Show/hide solution** button.  

```{asis, fold.chunk = TRUE}

Here is where a solution would be, but be sure to think about the problem on your own first!

```

(Careful: in your browser, the triangle for the Show/hide solution button might be close to the back button, so clicking on Show/hide might take you to the previous page.  To avoid this, click on the words **Show/hide**.)

<!-- ### Don't do what Donny Don't does {-} -->

<!-- Some of the examples in this book feature a character named Donny Don't. The moral of these examples is "Don't do what Donny Don't does". This is a [Simpson's](https://frinkiac.com/video/S05E08/0nvMY69o6o_U7BqeIzQ314al-SQ=.gif) [reference](https://simpsons.fandom.com/wiki/Boy-Scoutz_%27n_the_Hood). Donny represents a student who makes many of the mistakes commonly made by students studying probability.  The idea of these problems is for you to learn from the common mistakes that Donny makes, by identifying why he is wrong and by helping him understand and correct his mistakes.  (But be careful: sometimes Donny is right!) -->

<!-- ### Learning Objectives/Goals/Style??? (Better title) -->

<!-- - Don't skimp on rigorous definitions (RV is function defined on probspace) but deemphasize mathematical computation (counting and calculus) -->
<!-- - Emphasize simulation -->
<!-- - Visualize in lots of plots -->
<!-- - Start multivariate relationships early -->
<!-- - Rely on statistical literacy -->
<!-- - Active learning, workbook style -->








<!-- ### About this book {-} -->

<!-- This book was written using the [bookdown](https://bookdown.org/) package [@R-bookdown], which was built on top of R Markdown and [knitr](https://yihui.org/knitr/) [@xie2015].   -->




