--- 
title: "An Introduction to Bayesian Reasoning and Methods"
author: "Kevin Ross"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: 
  bookdown::gitbook:
    split_by: section
    config:
      toc:
        collapse: section
        scroll_highlight: yes
        before: null
        after: null
      toolbar:
        position: fixed
      edit : null
      download: null
      search: yes
      fontsettings:
        theme: white
        family: sans
        size: 2
      sharing:
        facebook: no
        twitter: no
        google: no
        linkedin: no
        weibo: no
        instapper: no
        vk: no
        all: ['facebook', 'google', 'twitter', 'linkedin', 'weibo', 'instapaper']
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bayesian-reasoning-and-methods
description: "This textbook presents an introduction to Bayesian reasoning and methods"
---

# Preface{-}

```{r include=FALSE}

library(knitr)
# library(learnr)
library(bookdown)
library(tidyverse)
library(ggplot2)
library(scales)
library(janitor)
library(viridis)

blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold", hjust = 0.5)
  )

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')


knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = TRUE)

options(scipen = 999, digits = 4)

set.seed(21234)


# knitr::opts_chunk$set(eval = FALSE)



hooks = knitr::knit_hooks$get()
hook_foldable = function(type) {
  force(type)
  function(x, options) {
    fold <- options[[paste0("fold.", type)]]
    part = hooks[[type]](x, options)
    
    # If fold is FALSE, don't fold
    if (isFALSE(fold)) return(paste0(part))

    # If fold is TRUE, fold
    else if (isTRUE(fold)) return(paste0(
      "<details><summary>", paste("Show/hide solution", " "), "</summary>\n\n",
      part,
      "\n\n</details>"
    ))
    
    # If fold is not specified, don't fold
    else return (paste0(part))

  }
}
knitr::knit_hooks$set(
  chunk = hook_foldable("chunk")
)


```


```{r, include = FALSE}
library(reticulate)
use_condaenv("Anaconda3")
# py_install("symbulate", pip = TRUE)
```


```{python, include = FALSE}
# These lines needed to show plots because of
# reticulate issue with matplotlib
# https://github.com/rstudio/rstudio/issues/4182

# import os
# os.environ['QT_QPA_PLATFORM_PLUGIN_PATH'] = 'C:/Users/kjross/Anaconda3/Library/plugins/platforms'

from symbulate import *
import matplotlib
import matplotlib.pyplot as plt
# from IPython.display import Markdown, display, HTML

import numpy as np

```




(ref:ddwddd) Don't do what Donny Don't does.


<!-- ### Why study Bayesian? {-} -->

<!-- ### Learn by doing {-} -->


Statistics is the science of learning from data.
Statistics involves

- Asking questions
- Formulating conjectures
- Designing studies
- Collecting data
- Wrangling data
- Summarizing data
- Visualizing data
- *Analyzing data*
- *Developing models*
- *Drawing conclusions*
- *Communicating results*



We will assume some familiarity with many of these aspects, and we will focus on the items in italics.
That is, we will focus on **statistical inference**, the process of using data analysis to draw conclusions about a population or process beyond the existing data.
"Traditional" hypothesis tests and confidence intervals that you are familiar with are components of *"frequestist" statistics*.
This book will introduce aspects of *"Bayesian" statistics*.
We will focus on analyzing data, developing models, drawing conclusions, and communicating results from a Bayesian perspective.
We will also discuss some similarities and differences between frequentist and Bayesian approaches, and some advantages and disadvantages of each approach.

We want to make clear from the start: Bayesian versus frequentist is NOT a question of "right versus wrong".
Both Bayesian and frequentist are valid approaches to statistical analyses, each with advantages and disadvantages.
We'll address some of the issues along the way.
But at no point in your career do you need to make a definitive decision to be a Bayesian or a frequentist; a good modern statistician is probably a [bit of both](https://tenor.com/view/star-lord-bit-of-both-chris-pratt-guardians-of-the-galaxy-peter-quill-gif-11821953).


While our focus will be on statistical inference, remember that the other parts of Statistics are equally important, if not more important.
In particular, any statistical analysis is only as good as the data upon which it is based.


The exercises in this book are used to both motivate new topics and to help you practice your understanding of the material.  You should attempt the exercises on your own before reading the solutions.  To encourage you to do so, the solutions have been hidden.  You can reveal the solution by clicking on the **Show/hide solution** button.  

```{asis, fold.chunk = TRUE}

Here is where a solution would be, but be sure to think about the problem on your own first!

```

(Careful: in your browser, the triangle for the Show/hide solution button might be close to the back button, so clicking on Show/hide might take you to the previous page.  To avoid this, click on the words **Show/hide**.)

<!-- ### Don't do what Donny Don't does {-} -->

<!-- Some of the examples in this book feature a character named Donny Don't. The moral of these examples is "Don't do what Donny Don't does". This is a [Simpson's](https://frinkiac.com/video/S05E08/0nvMY69o6o_U7BqeIzQ314al-SQ=.gif) [reference](https://simpsons.fandom.com/wiki/Boy-Scoutz_%27n_the_Hood). Donny represents a student who makes many of the mistakes commonly made by students studying probability.  The idea of these problems is for you to learn from the common mistakes that Donny makes, by identifying why he is wrong and by helping him understand and correct his mistakes.  (But be careful: sometimes Donny is right!) -->

<!-- ### Learning Objectives/Goals/Style??? (Better title) -->

<!-- - Don't skimp on rigorous definitions (RV is function defined on probspace) but deemphasize mathematical computation (counting and calculus) -->
<!-- - Emphasize simulation -->
<!-- - Visualize in lots of plots -->
<!-- - Start multivariate relationships early -->
<!-- - Rely on statistical literacy -->
<!-- - Active learning, workbook style -->








<!-- ### About this book {-} -->

<!-- This book was written using the [bookdown](https://bookdown.org/) package [@R-bookdown], which was built on top of R Markdown and [knitr](https://yihui.org/knitr/) [@xie2015].   -->




